---
title: "03 Troubleshooting example 2"
author: "Antonia Netzl (an604@cam.ac.uk)"
date: "2025-07-01"
output: html_document
---

```{r setup, include=FALSE}
rm(list = ls()) # clear environment at the beginning of each session
knitr::opts_chunk$set(echo = TRUE,
                      fig.width=8, fig.height=6,
                      fig.align = "center")
```

```{r load_packages, echo=FALSE}
# Package names
packages <- c("Racmacs", "tidyverse")

# Install packages not yet installed
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

invisible(lapply(packages, library, character.only = TRUE))

#set ggplot2 theme
theme_set(theme_bw() + 
            theme(strip.background.x = element_blank(),
                  axis.text.x = element_text(angle = 45, hjust = 1)))
```

## Troubleshooting example 2

This example contains two maps, example map 2a and example map 2b. Working code is given in [`03_troubleshooting_ex4_solution.Rmd`](./code/03_troubleshooting_ex4_solution.Rmd).

### Troubleshooting example 2a

Load the map from `data/maps/03_troubleshooting_map2a.ace` and explore:

```{r}

```
### Troubleshooting example 2b

Load the map from `data/maps/03_troubleshooting_map2b.ace` and explore:

```{r}

```


