---
title: "03 Troubleshooting example 3"
author: "Antonia Netzl (an604@cam.ac.uk)"
date: "2025-06-30"
output: html_document
---

```{r setup, include=FALSE}
rm(list = ls()) # clear environment at the beginning of each session
knitr::opts_chunk$set(echo = TRUE,
                      fig.width=8, fig.height=6,
                      fig.align = "center")
```

## Troubleshooting example 3

This is a real-world example, previously published in [Rössler, et al. Cell Reports (2025)](https://doi.org/10.1016/j.celrep.2024.115140.). Working code is given in [`03_troubleshooting_ex5_solution.Rmd`](./code/03_troubleshooting_ex5_solution.Rmd).


```{r load_packages, echo=FALSE}
# Package names
packages <- c("Racmacs", "knitr", "tidyverse")

# Install packages not yet installed
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

invisible(lapply(packages, library, character.only = TRUE))

#set ggplot2 theme
theme_set(theme_bw() + 
            theme(strip.background.x = element_blank(),
                  axis.text.x = element_text(angle = 45, hjust = 1)))

```

Load the map from `data/maps/03_troubleshooting_map3.ace` . The original map by Rössler, et al., can be downloaded [here](https://github.com/acorg/roessler_netzl_et_al2024/blob/main/data/maps/map_threshold20_all_ags_singleTP_woXBBBQ11conv_woXBB15conv_alpha_adj.ace). 

Explore the map with the tools from the workshop and see what you can find. You can go to `code/03_troubleshooting_example3.Rmd` for hints and one example 
analysis pathway.

Have fun!

```{r}

```